### 2023-11-29 draft_jss/draft_jss.Rmd
update manuscript

```{r covidintro, fig.cap="Hospitalizations, IPTCC and positive PCR of Bordeaux University Hospital.", echo = FALSE}
dfOutcome %>%
  tidyr::pivot_longer(cols = c("hosp", "Positive", "Tested")) %>%
  ggplot2::ggplot(mapping = aes(x = date, y = value)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y") +
  theme_bw() +
  geom_vline(mapping = aes(color = "train-test sets",
                           xintercept = traintest_date)) +
  labs(color = "") +
  theme(legend.position = "bottom")
```

@@

```{r nodeactivationbad, fig.cap="Node activation over time."}
plot(reservoir_state)
```

Numerous nodes within the system exhibit a consistent equilibrium state. The challenge arises when the ridge output layer attempts to acquire knowledge from these nodes, as they do not convey meaningful information. This issue can be attributed to the disparate scales of the features. To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of `-1` to `1`.

```{r}
stand_max <- function(x) return(x/max(x))
# scaled features
Xstand <- dfOutcome %>%
  filter(date < traintest_date) %>%
  select(hosp, Positive, Tested) %>%
  mutate_all(.funs = stand_max) %>%
  as.matrix() %>%
  as.array()
```

@@

```{r nodeactivationgood, fig.cap="Node acivation over time. Scaled features"}
# feed them to the reservoir
reservoir_state_stand <- predict_seq(node = reservoir,
                                     X = Xstand,
                                     reset = TRUE)
# plot the output
plot(reservoir_state_stand)
```

@@

```{r fig.cap="Forecast"}
dfOutcome %>%
  mutate(pred = vec_pred) %>%
  na.omit() %>%
  ggplot(mapping = aes(x = outcomeDate)) +
  geom_line(mapping = aes(y = outcome,
                          color = "observed")) +
  geom_line(mapping = aes(y = pred,
                          color = "forecast")) +
  geom_vline(mapping = aes(color = "train-test sets",
                           xintercept = traintest_date)) +
  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
  theme_bw() +
  labs(color = "", x = "Date", y = "Hospitalizations")
```

We observe that the model forecast is not fully accurate, both on the test set and the train set. In that case, one option could be to reduce ridge penalization to fit more closely the data, the optimization of ridge hyperparameter will be discussed at section [Study case](#studycase). Another possibility is to ease the learning of the algorithm by forecasting the variation of the hospitalization instead of the number of hospitalized patients. For that step, we will learn on the `outcome_deriv` contained in `yTrain_variation` data which is defined outcome as `outcome_deriv = outcome - hosp`.

```{r fig.cap="Covid-19 hospitalizations forecast. The model is either trained to forecast the number of hospitalizations (denoted Raw) or the variation of the hospitalizations compared to current level of hospitalisation (denoted Variation)"}
## Fit reservoir on outcome variation instead of raw outcome
fit2 <- reservoirnet::reservoirR_fit(node = model,
                                     X = lsdf$xTrain,
                                     Y = lsdf$yTrain_variation,
                                     warmup = 30,
                                     reset = TRUE)
## Get the forecast on the test set
vec_pred2_variation <- reservoirnet::predict_seq(node = fit2$fit,
                                                 X = lsdf$xTest,
                                                 reset = TRUE)
## Transform the outome variation forecast into hospitalisation forecast
vec_pred2 <- vec_pred2_variation + xTest$hosp
## Plot the results
dfOutcome %>%
  mutate(Raw = vec_pred,
         Variation = vec_pred2) %>%
  tidyr::pivot_longer(cols = c(Raw, Variation),
                      names_to = "Outcome_type",
                      values_to = "Forecast") %>%
  na.omit() %>%
  ggplot(mapping = aes(x = outcomeDate)) +
  geom_line(mapping = aes(y = outcome,
                          color = "observed")) +
  geom_line(mapping = aes(y = Forecast,
                          color = "Forecast")) +
  geom_vline(mapping = aes(color = "train-test sets",
                           xintercept = traintest_date)) +
  facet_wrap(Outcome_type ~ .,
             labeller = label_bquote(cols = "Outcome" : .(Outcome_type))) +
  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
  theme_minimal() +
  labs(color = "", x = "Date", y = "Hospitalizations")
```

We observe an improvement of the forecast compared to the forecast of hospitalization previously discussed. From there, many improvement can be implemented including : frequent update of the model, leverage additional features and hyperparameter optimization. This notions will be discussed in the [Study case section](#studycase).

@@

ggplot(dfplotseqtoseq, mapping = aes(x = Time,
                                     y = pred_speaker,
                                     fill = prediction)) +
  geom_tile() +
  facet_wrap(uterrance ~ speaker,
             labeller = label_bquote(cols = "speaker" : .(speaker)),
             ncol = 2) +
  theme_minimal() +
  labs(y = 'Predicted speaker',
       fill = "Prediction score")
```

@@

First derivative over the last $7$ days were computed to enrich model information. To take into account measurement error and daily noise variation, data were smoothed using a local polynomial regression with a span of 21 days.

@@

@@

Because of the randomness of the reservoir, we took the median forecast of 20 reservoir on the train set and 40 reservoir on the test set. In addition, because covid-19 hospitalization is a non-stationary process, models were re-trained everyday using all previous days. To ease computation burden, only one day over two was used to find hyperparameters on the training set.

@@

```{r ucPresentData, echo = FALSE, fig.cap = "Covid-19 epidemic at BUH. Outcome of interest is presented in orange. Model is trained to forecast Outcome curve which corresonds to the difference between Hospitalisatiosn at 14 days and current hospitalisations. Other features are represented in darkblue.", fig.height=8}
ls_use_case$plot_present_data
```

@@

Figure \@ref(fig:ucHyperparam) shows the hyperparameter optimisation using random search for the different RC architectures. We note that the multiple input scaling setting with input layer connected to output layer showed poor performance on the training set which might be due to an unextensive search of hyperparameter regarding the better performance of the common input scaling with connected input layer which is a specific configuration where all input scaling are equal.

As expected, we observe that the optimal leaking rate is close to 1 for all RC which is coherent with the short term dynamic of covid-19 epidemic. Trend for other hyperparameter are less clear even though best hyperparameter sets were close for RC with common input scaling.

@@

Figure \@ref(fig:ucHyperparamMultipleIS) shows the hyperparameter search for RC with multiple input scaling and connected input layer. We observe that the random search tend to favor low importance given to vaccine and high importance given to derivative of hospitalisations and derivative of positive RT-PCR in the elderly. The rest of the feature do not show clear pattern. The lack of importance of number of people with one dose of vaccine in this setting was discussed by @ferte_benefit_2022 and related to the lack of vaccinated people in the training set as depicted at figure \@ref(fig:ucPresentData).

@@

Table \@ref(tab:performance) shows the performance on the test set. Best model according to all metrics was RC with common input scaling and no connection between input and output layers. Having one input scaling per feature did not improve the model which might be due to low generalisability of the hyperparameter of the training set to the test set due to non-stationarity. Additionaly, connecting input layer to output layer did not improve the model, possibly because it add more dimension to the problem while the information provided is already inside the reservoir. We observe that elastic-net did not perform well and stay close to the baseline prediction. However, when the penalisation is updated each day, the peformance improve. This is emphasize the problem of the transferrability of the hyperparameters to which multiple input scaling models seem more sensitive as they have more hyperparameters. For instance, vaccine will keep a low weight whereas its low initial importance quickly increase to mitigate the effect of the high RT-PCR increase on the hospitalisation dynamic.

Figure \@ref(fig:forecastwithorwithoutUpdate) shows the forecast of the different models. We note that models struggle to accurately forecast slope shifts. For instance, summer 2021 initial increase is missed by all models, summer decrease is partially anticipated by RC with common IS and no input layer connection, winter 2021 increase is anticipated by all models but they tend to overestimate it because of the rise of vaccine effect.

@@

Figures \@ref(fig:ucAggregationSpread) and figure \@ref(fig:ucAggregationNb) show 40 model forecast for each of the best hyperparameter set of the different reservoir architecture (i.e the same set of hyperparameters as used for test set forecast). It means at each day, the same set of hyperparameter is kept but a new reservoir is generated. Due to the internal random connection of the reservoir, it might result in forecast stochasticity. We observe that models with ridge penalty higher than 1 (i.e all models except multiple IS with input connection) tend to have more stable forecast but that the addition of input layer connection provided some extreme forecasted value. This extreme value phenomena is relatively rare because it is corrected when the number of model for aggregation is above 15.

```{r ucAggregationSpread, echo = FALSE, fig.cap = "Individual forecast of 40 repetition of the best hyperparameter set for the different RC configuration. Forecast value above 200 were set to 200.", fig.height=6}
ls_use_case$plot_before_aggregation_forecast
```

```{r ucAggregationNb, echo = FALSE, fig.cap = "It shows the impact of aggregating several RC with the best set of HP for each RC configuration. It has small interest when the input layer was not connected to output layer but significantly improves the performance otherwise.", fig.height=6}
ls_use_case$plot_aggregation
```

@@

This study represents a novel contribution to epidemic forecasting utilizing reservoir computing. Notably, previous literature predominantly focused on simpler problems characterized by fewer input features or shorter evaluation periods. Our findings underscore the potential of this approach for future epidemics, suggesting its potential to surpass more traditional epidemiological tools while maintaining a lightweight model structure compared to other recurrent neural networks (RNNs).

@@

Given the substantial number of hyperparameters involved, we acknowledge that random search may not be the most efficient optimization algorithm. We have retained this approach for the sake of simplicity in this tutorial paper; however, meta-heuristic approaches, particularly those utilizing evolutionary algorithms, may prove more efficient, especially when employing multiple input scaling.

>>>>

```{r covidintro, fig.cap="Hospitalizations, IPTCC and positive PCR of Bordeaux University Hospital.", echo = FALSE}
dfOutcome %>%
  tidyr::pivot_longer(cols = c("hosp", "Positive", "Tested")) %>%
  ggplot2::ggplot(mapping = aes(x = date, y = value)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y") +
  theme_minimal() +
  geom_vline(mapping = aes(color = "train-test sets",
                           xintercept = traintest_date)) +
  labs(color = "") +
  theme(legend.position = "bottom")
```

@@

```{r nodeactivationbad, fig.cap="20 random nodes activation over time."}
plot(reservoir_state)
```

Numerous nodes within the system exhibit a consistent equilibrium state. The challenge arises when the ridge output layer attempts to acquire knowledge from these nodes, as they do not convey meaningful information. This issue can be attributed to the disparate scales of the features. To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of `0` to `1` by dividing by the maximum of the absolute value.

```{r}
stand_max <- function(x) return(x/max(abs(x)))
# scaled features
Xstand <- dfOutcome %>%
  filter(date < traintest_date) %>%
  select(hosp, Positive, Tested) %>%
  mutate_all(.funs = stand_max) %>%
  as.matrix() %>%
  as.array()
```

@@

```{r nodeactivationgood, fig.cap="20 random node acivation over time. Scaled features."}
# feed them to the reservoir
reservoir_state_stand <- predict_seq(node = reservoir,
                                     X = Xstand,
                                     reset = TRUE)
# plot the output
plot(reservoir_state_stand)
```

@@

```{r fig.cap="Forecast"}
dfOutcome %>%
  mutate(pred = vec_pred) %>%
  na.omit() %>%
  ggplot(mapping = aes(x = outcomeDate)) +
  geom_line(mapping = aes(y = outcome,
                          color = "observed")) +
  geom_line(mapping = aes(y = pred,
                          color = "forecast")) +
  geom_vline(mapping = aes(color = "train-test sets",
                           xintercept = traintest_date)) +
  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
  theme_minimal() +
  labs(color = "", x = "Date", y = "Hospitalizations")
```

We observe that the model forecast is not fully accurate, both on the test set and the train set. In that case, one option could be to reduce ridge penalization to fit more closely the data, the optimization of ridge hyperparameter will be discussed at section [Study case](#studycase). Another possibility is to ease the learning of the algorithm by forecasting the variation of the hospitalization instead of the number of hospitalized patients. For that step, we will learn on the `outcome_deriv` contained in `yTrain_variation` data which is defined outcome as `outcome_deriv = outcome - hosp`. As depicted at figure \@ref(fig:figForecastDerivvsRaw), this strategy improved the model forecast.

```{r figForecastDerivvsRaw, fig.cap="Covid-19 hospitalizations forecast. The model is either trained to forecast the number of hospitalizations (denoted Raw) or the variation of the hospitalizations compared to current level of hospitalisation (denoted Variation)"}
## Fit reservoir on outcome variation instead of raw outcome
fit2 <- reservoirnet::reservoirR_fit(node = model,
                                     X = lsdf$xTrain,
                                     Y = lsdf$yTrain_variation,
                                     warmup = 30,
                                     reset = TRUE)
## Get the forecast on the test set
vec_pred2_variation <- reservoirnet::predict_seq(node = fit2$fit,
                                                 X = lsdf$xTest,
                                                 reset = TRUE)
## Transform the outome variation forecast into hospitalisation forecast
vec_pred2 <- vec_pred2_variation + xTest$hosp
## Plot the results
dfOutcome %>%
  mutate(Raw = vec_pred,
         Variation = vec_pred2) %>%
  tidyr::pivot_longer(cols = c(Raw, Variation),
                      names_to = "Outcome_type",
                      values_to = "Forecast") %>%
  na.omit() %>%
  ggplot(mapping = aes(x = outcomeDate)) +
  geom_line(mapping = aes(y = outcome,
                          color = "observed")) +
  geom_line(mapping = aes(y = Forecast,
                          color = "Forecast")) +
  geom_vline(mapping = aes(color = "train-test sets",
                           xintercept = traintest_date)) +
  facet_wrap(Outcome_type ~ .,
             labeller = label_bquote(cols = "Outcome" : .(Outcome_type))) +
  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
  theme_minimal() +
  labs(color = "", x = "Date", y = "Hospitalizations")
```

@@

ggplot(dfplotseqtoseq, mapping = aes(x = Time,
                                     y = pred_speaker,
                                     fill = prediction)) +
  geom_tile() +
  facet_wrap(uterrance ~ speaker,
             labeller = label_bquote(cols = "speaker" : .(speaker)),
             ncol = 2) +
  scale_fill_gradient2(low = "#8ECAE6", high = "#FB8500", mid = "#023047",
                       midpoint = 0) +
  theme_minimal() +
  labs(y = 'Predicted speaker',
       fill = "Prediction score")
```

@@

First derivative over the last $7$ days were computed to enrich model information. To take into account measurement error and daily noise variation, data were smoothed using a local polynomial regression with a span of 21 days. As previously described, input features were scaled between -1 and 1 by dividing the observed value by the maximum of the absolute value of the given input feature.

@@

Because the outcome is obfuscated below 10 hospitalizations for privacy reason, we set both the outcome and the forecast to 10 when the observed value was 0 or the forecasted value was below 10 when evaluating the model performance.

@@

Because of the randomness of the reservoir, we took the median forecast of 10 reservoir on the train set to evaluate the performance of a given hyperparameter set. On the test set we aggregated the forecast of 40 reservoir, each of them having one of the 40 best hyperparameter sets found on the train set. In addition, because covid-19 hospitalization is a non-stationary process, models were re-trained everyday using all previous days. To ease computation burden, only one day over two was used to find hyperparameters on the training set.

@@

```{r ucPresentData, echo = FALSE, fig.cap = "Covid-19 epidemic at BUH. Outcome of interest is presented in orange. Model is trained to forecast Outcome curve which corresonds to the difference between Hospitalisatiosn at 14 days and current hospitalisations. Other features are scaled (divide by the maximum of the feature) represented in darkblue.", fig.height=8}
ls_use_case$plot_present_data
```

@@

Figure \@ref(fig:ucHyperparam) shows the hyperparameter optimisation using random search for the different RC architectures. We observe that model with multiple input scaling achieved better performance on the train set compared to model with single input scaling which is expected as they can adapt more closely to the data thanks to specific input scaling for each feature.

As expected, we observe that the optimal leaking rate is above 1e-2 for all RC which is coherent with the short term dynamic of covid-19 epidemic. Trend for other hyperparameter are less clear even though best hyperparameter sets were close for RC with common input scaling and for RC with multiple input scaling.

@@

Figure \@ref(fig:ucHyperparamMultipleIS) shows the hyperparameter search for RC with multiple input scaling and connected input layer. We observe that the random search tend to favor high importance given to derivative of positive RT-PCR (including the elderly) and the derivative of IPTCC. The rest of the feature do not show clear pattern.

@@

Table \@ref(tab:performance) shows the performance on the test set. Best model according to all metrics was RC with common input scaling and connection between input and output layers. Having one input scaling per feature did not improve the model which might be due to low generalisability of the hyperparameter of the training set to the test set due to non-stationarity. Additionaly, connecting input layer to output layer improved the model forecast.

We observe that elastic-net did not perform well and stayed close to the baseline prediction. However, when the penalisation is updated each day, the peformance improve. This is emphasize the problem of the transferrability of the hyperparameters to which multiple input scaling models seem more sensitive as they have more hyperparameters.

Figure \@ref(fig:forecastwithorwithoutUpdate) shows the forecast of the different models. We note that models struggle to accurately forecast slope shifts. For instance, summer 2021 initial increase is partially predicted by all models but its decrease is not well predicted. Winter 2021 increase is anticipated by all models but they tend to overestimate it because of the rise of vaccine effect.

@@

Figures \@ref(fig:ucAggregationSpread) show the individual forecast for the 40 best set of hyperparameters of each RC architecture. Due to the internal random connection of the reservoir, we observe forecast stochasticity and relying on only one forecast is unreliable. We explored the number of model needed at figure \@ref(fig:ucAggregationNb) which shows that after 10 models, forecast is stable and even 5 models for the simpler model with common input scaling which rely on less hyperparamters.

```{r ucAggregationSpread, echo = FALSE, fig.cap = "Individual forecast the 40 best hyperparameter sets for the different RC configuration. Forecast value above 200 were set to 200 for clarity.", fig.height=6}
ls_use_case$plot_before_aggregation_forecast
```

```{r ucAggregationNb, echo = FALSE, fig.cap = "Mean absolute error depending on the number of aggregated reservoir.", fig.height=6}
ls_use_case$plot_aggregation
```

@@

@@

Given the substantial number of hyperparameters involved, we acknowledge that random search may not be the most efficient optimization algorithm. We have retained this approach for the sake of simplicity in this tutorial paper; however, meta-heuristic approaches, particularly those utilizing evolutionary algorithms, may prove more efficient, especially when employing multiple input scaling [@bala_applications_2018].

This study represents a novel contribution to epidemic forecasting utilizing reservoir computing. Notably, previous literature predominantly focused on simpler problems characterized by fewer input features or shorter evaluation periods [@ghosh_reservoir_2021, @liu_nanophotonic_2023, @ray_optimized_2021]. Our findings underscore the potential of this approach for future epidemics, suggesting its potential to surpass more traditional epidemiological tools while maintaining a lightweight model structure compared to other recurrent neural networks (RNNs).
